Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-15 draft.tex
revised version from Sylvia

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.

@@

\emph{Hypotheses generation} amplified by concurrently analyzing heterogenous variables at once using correlation measures.

@@

\cite{Turkay} derive descriptive data metrics from image data.

@@

This approach aims to trigger \emph{hypotheses generation} by outlining tendencies between these variables.

@@

We incorporate hierarchical agglomerative clustering to derive shape groups, yielding average shape groups and outliers.

@@

\cite{Zhang} analyze subject groups in a web-based linked view system.

@@

They define a cohort as variable-divided subject group, differing from the epidemiological understanding of the term.

@@

Both use decision trees for their analyses.

@@

\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in tumorous tissue.

@@

To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological variables alone are not sufficient for tumor type classification.

@@

\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) using decision trees.

@@

\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.

@@

	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.

@@

	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes).
	The \emph{curvature angle} ($\alpha$) is calculated using the top, bottom and middle point.

@@

Larger people have also a longer spine with a straighter shape.

@@

We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at \url{ivapp15.dnsalias.com}).

@@

The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.

@@

A Generalized Pairs Plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.

@@

No correlations with back pain are present as it can be seen in this plot.

@@

The method uses \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{polychoric} correlations for two categorical variables.

@@

These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
%%
% \section{\uppercase{Evaluation of Decision Trees}}
\section{\uppercase{Interactive Decision Tree Visualization}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
%%
%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
%The nature of this analysis is more explorative as it includes all 
%%
Decision trees are used to create predictive models.
%%
These models are build w.r.t. all input variables and capture more complex relationships than correlation coefficients.
%%
Leafs of a decision tree represent class labels, branches represent variable conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
%They work with numerical as well as categorical data.
%%
Readability is only granted for small trees.
%%
%Large, complex trees are difficult to trace and to understand.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%
%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%\subsection{Visualizing Decision Trees}

@@

The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
This makes these results less reliable.

@@

Metric variables, such as \emph{body size} are discretized using their quantiles.

@@

This allows to assess the influences of a variable to the classification process.

@@

The left scatter plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.

@@

The right scatter plot displays the dummy variables derived from the original variables.

@@

We only display variables with an error below $0.5$, results above this threshold are inaccurate.

@@

The interactive plot (see supplemental material) has clickable data points, displaying corresponding decision tree in a tool tip.

@@

Almost all medicated subjects were correctly classified ($796$ of $1,058$).

@@

The classifier neglects the other group and yields an error below $0.5$.

@@

Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$

@@

They are an addition to the standard epidemiological workflow as they highlight new possibly unknown relationships.

@@

Dented vertebras hint a pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.

>>>>

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically train a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

For example, the incidence rate of many diseases, such as different cancer types increases with age \cite{Fletcher}.
%Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.

@@

\emph{Hypotheses generation} was amplified by concurrently analyzing heterogenous variables at once using correlation measures.

@@

\cite{Turkay} derived descriptive data metrics from image data.

@@

This approach aimed to trigger \emph{hypotheses generation} by outlining tendencies between these variables.

@@

We incorporated hierarchical agglomerative clustering to derive shape groups, yielding average shape groups and outliers.

@@

\cite{Zhang} analyzed subject groups in a web-based linked view system.

@@

They defined a cohort as variable-divided subject group, differing from the epidemiological understanding of the term.

@@

Both used decision trees for their analyses.

@@

\cite{Glasser2013} used variables derived from DCE-MRI data capturing the perfusion in tumorous tissue.

@@

To classify breast tumors, they trained a decision-tree classifier, concluding that the extracted kinetic and morphological variables alone are not sufficient for tumor type classification.

@@

\cite{Niemann2014} assessed risk factors for hepatic steatosis (fatty liver disease) using decision trees.

@@

\noindent Our work is based on a data set compiled by Epidemiologists with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
%\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.

@@

	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure}, and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.

@@

	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes) and the \emph{curvature angle} ($\alpha$). %is calculated using the top, bottom and middle point.

@@

Larger people also have a longer spine with a straighter shape.

@@

We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a principal component analysis (see supplementary material at \url{ivapp15.dnsalias.com}).

@@

The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub-plots.

@@

A \texttt{GPLOM} of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.

@@

No correlations with back pain can be identified in this plot.

@@

The method uses the \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable and \emph{polychoric} correlations for two categorical variables.

@@

These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
%%
% \section{\uppercase{Evaluation of Decision Trees}}
\section{\uppercase{Interactive Decision Tree Visualization}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
%%
%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
%The nature of this analysis is more explorative as it includes all 
%%
Decision trees are used to create predictive models.
%%
These models are built w.r.t. all input variables and capture more complex relationships than correlation coefficients.
%%
Leafs of a decision tree represent class labels, branches represent variable conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
%They work with numerical as well as categorical data.
%%
%Readability is only granted for small trees.
%%
%Large, complex trees are difficult to trace and to understand.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%
%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%\subsection{Visualizing Decision Trees}

@@

The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size, making these results less reliable.

@@

Metric variables, such as \emph{body size}, are discretized using their quantiles.

@@

This allows to assess the influence of a variable to the classification process.

@@

The upper scatter plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.

@@

The lower scatter plot displays the dummy variables derived from the original variables.

@@

We only display variables with an error below $0.5$, results above this threshold are dismissed.

@@

The interactive plot (see supplemental material) has clickable data points, displaying the corresponding decision tree in a tool tip.

@@

A high share of medicated subjects were correctly classified ($796$ of $1,058$).

@@

The classifier neglects the other groups and yields an error below $0.5$.

@@

This result also imposes overfitting.
Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$.

@@

They are an addition to the standard epidemiological workflow as they highlight new and possibly unknown relationships.

@@

Dented vertebrae indicate pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.
